Illuminating a Dark Lens : A Type Ia Supernova \\ Magnified by Galaxy Cluster Abell 2744

### 2015-04-22 TABLE/magnifications.tex
incorporating minor comments on draft, post redshift change   * notably: updated references and minor rearrangement of section 5  and new text in 6.6

CATS(v2)     & \nodata  & 3.42  &   3.27$-$3.58 &   HFF\tablenotemark{d}; \citealt{Jauzac:2014} & Updated version of the CATSv1 model, adds 33 new multiply-imaged galaxies, for a total of 159 individual lensed images.\\

>>>>

CATS(v2)     & \nodata  & 3.42  &   3.27$-$3.58 &   HFF\tablenotemark{d}; \citealt{Jauzac:2014c} & Updated version of the CATSv1 model, adds 33 new multiply-imaged galaxies, for a total of 159 individual lensed images.\\

### 2015-04-22 snTomas.tex
incorporating minor comments on draft, post redshift change   * notably: updated references and minor rearrangement of section 5  and new text in 6.6

\begin{figure*}
\begin{center}
%\includegraphics[width=\textwidth]{FIG/discovery_image}
\includegraphics[width=\textwidth]{FIG/discovery_image_lowres}
\caption{  \label{fig:DiscoveryImage} 
SN \tomas\ in the Abell 2744 field.  The left panel shows a
UV/Optical/IR color composite image constructed from all available HST
imaging of the Abell 2744 cluster field.  The inset panels on the right
show F814W imaging of the immediate vicinity of SN \tomas,
approximately 40\arcsec\ from the center of the cluster. The top panel
shows the template image, combining all data prior to the SN
appearance.  Labeled ellipses mark the nearest galaxies and their
redshift constraints: a photometric redshift for the most likely host
galaxy is marked in blue, and the spectroscopic redshift for a
background galaxy is given in yellow. The bottom panel is constructed
from all HFF F814W imaging taken while the SN was detectable, and
marks the SN location with an arrow.  (Left panel image credit: NASA,
ESA, and J. Lotz, M. Mountain, A. Koekemoer, and the HFF Team (STScI))
}
\end{center}
\end{figure*}

@@

To refine the redshift and phase constraints on \tomas, we next fit
the spectrum with a custom spectral matching program that employs a
library of Type Ia SN SEDs.  This library is similar to the Type Ia
spectral set used by SNID, but also includes more recent SNe with
well-observed spectral time series that extend to rest-frame UV
wavelengths (e.g. SN 2011fe and 2014J).  We first use an approach
similar to the SNID algorithm: warping the pseudo-continuum of each
template spectrum by dividing out a 3$^{\rm rd}$-order polynomial to
match the observed SED of SN \tomas.  \changeG{This approach will find
templates that have similar abundances and photospheric
velocities. With the redshift fixed at $z=1.3457$ we find the best fit
is from the normal Type Ia SN 2014J, with a $\chi^2$ per degree of
freedom $\nu$ of $\chi^2/\nu=121.8/89$ shown in the top panel of
Figure~\ref{fig:SpecFit}.  Allowing the redshift as an additional free
parameter, we still find results that are consistent with the SNID
fits and $<2\sigma$ from the host galaxy redshift: $z=1.31\pm0.02$ and
a phase of $0\pm3$ rest-frame days. The best-fitting spectral template
in this case is the normal Type Ia SN 2012cg at $z=1.31$ with
$\chi^2/\nu=98.5/88$ (second panel of Figure~\ref{fig:SpecFit}).}

\changeG{
Next, we repeat the fitting, but without any warping of the templates
to account for differences in the continuum shape.  In this iteration
we only allow each template SED to be scaled in flux coherently at all
wavelengths, so the fits are more sensitive to the overall color of
the SED. Fixing the redshift to $z=1.3457$, we find the best match is
from the normal Type Ia SN 2011fe, though the fit is quite poor, with
$\chi^2/\nu=313.2/91$ (third panel of Figure~\ref{fig:SpecFit}).  When
the redshift is allowed as a free parameter the \tomas\ SED is still
matched best by a SN 2011fe template, now at redshift $z=1.31$ with
$\chi^2/\nu=240.8/90$ (fourth panel of Figure~\ref{fig:SpecFit}).  }

Without the continuum warping, an alternative fit also arises: the
fast-declining (91bg-like) Type Ia SN 2011iv at
$z=0.98\pm0.01$. Formally, this match provides a slightly better fit
to the unwarped \tomas\ spectrum \changeG{($\chi^2/\nu=220.6/90$,
bottom panel of Figure~\ref{fig:SpecFit}),} although the fit is
notably poorer at $\sim8700$\AA\ where the most significant absorption
feature is found. Furthermore, a redshift $z\sim1$ is at odds with the
spectroscopic redshift of the nearest galaxy ($z=1.3457$), and we will
see in the following section that the photometric data is also
incompatible with a Type Ia SN at $z\sim1.0$.

@@

\section{Light Curve Fitting}
\label{sec:LightCurveFitting}

With the type and redshift securely defined as a normal
Type Ia SN at $z=1.3457$, we now turn to fitting the light curve with
Type Ia templates to measure the distance modulus.  Here we use two
independent light curve fitters: the SALT2 model described
above and the MLCS2k2 model \citep{Jha:2007}.

@@

With both fitters we find light curve shape and color parameters that
are fully consistent with a normal Type Ia SN.  For SALT2, with the
\change{redshift fixed at $z=1.3457$, we find a light curve shape
parameter of $x_1=0.135\pm0.199$ and a color parameter of
$c=-0.127\pm0.025$, yielding a $\chi^2$ value of 45.0 for 36 degrees
of freedom, $\nu$.  With the MLCS2k2 fitter the best-fit shape
parameter is $\Delta=-0.082\pm0.070$ and the color term is
$A_{\rm V}=0.011\pm0.025$, giving $\chi^2/\nu=22.8/36$.}

\subsection{Distance Modulus}
\label{sec:DistanceModulus}

As in \citealt{Patel:2014}, we derive a distance modulus (dm)\footnote{We use 'dm' to
indicate the distance modulus to avoid confusion, reserving the symbol
$\mu$ to refer to the lensing magnification. This 'dm' is a standard
distance modulus, defined as ${\rm dm}=5\log_{10}d_L+25$, where $d_L$
is the luminosity distance in Mpc.} while ignoring the
effects of gravitational lensing, so that we can later use it to
derive a direct measurement of the magnification.  \change{The MLCS2k2 fitter returns $\rm{dm}_{\rm MLCS2k2}$ directly, as it is defined as one of the free parameters in the model. To derive dm from} the SALT2 fit
we use

@@

Final values for the SN distance modulus are shown in
Table~\ref{tab:MeasuredMagnification}, adopting the spectroscopic
redshift ($z=1.3457$).  In addition to modifying the distance modulus
uncertainty for SN \tomas, we also add $\sigma_{\rm int}=0.08$ mag in
quadrature to the uncertainty for every SN in the control sample. Note
however that the uncertainty on the control sample value at $z=1.3457$
is only 0.06 mag, smaller than the intrinsic dispersion because it
reflects only our measurement error on the mean distance modulus of
the population.  The distance moduli derived from the SALT2 and
MLCS2k2 light curve fitters are fully consistent within the
uncertainties.

\subsection{Host Galaxy Mass Correction}
\label{sec:HostGalaxyMassCorrection}

@@

The physical mechanism that drives the mass step is not yet
understood, but may be related to the metallicity or age of the SN
progenitor systems.  In either case, the significance of this effect
should decrease with redshift, as metal-rich passive galaxies become
much less common at $z>1$ \change{\citep[see
e.g.][]{Rigault:2013,Childress:2014b}.}  Indeed, when the size of the
mass step correction is allowed to vary with redshift, there is no
significant evidence that a non-zero mass step is required at
$z>1$ \citep{Rigault:2013,Shafer:2014,Betoule:2014}.  Given the
absence of a clear physical model and the lack of empirical support
for a high-$z$ mass step, we do not apply any correction to SN \tomas\
or the control sample.

\subsection{Magnification}
\label{sec:Magnification}

To measure the lensing magnification, we would like to avoid
introducing systematic uncertainties inherent to any assumed
cosmological model \citep[e.g.][]{Nordin:2014}.  To that end, we
follow \citealt{Patel:2014} and define the magnification by comparing
the measured distance modulus of SN \tomas\ against an average
distance modulus derived from a ``control sample'' of unlensed Type Ia
SN at similar redshift.  This allows us to make only the minimal
assumption that the redshift-distance relationship for Type Ia SN is
smooth and approximately linear over a small redshift span, which
should be true for any plausible cosmological model.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{FIG/snTomas_magnifications}
\caption{ 
Comparison of the observed lensing magnification to predictions from
lens models. The vertical blue line shows the constraints from
SN \tomas\ derived in Section~\ref{sec:Magnification} using the
MLCS2k2 fitter, with a shaded region marking the total
uncertainty. Markers with horizontal error bars show the median
magnification and 68\%\ confidence region from each of the 14 lensing
models.  Circles indicate models that use only strong-lensing
constraints, while diamonds denote those that also incorporate
weak-lensing measurements.  Models shown with open markers were
constructed using only data available before the start of the Frontier
Fields observations.  Filled markers indicate models that use
additional input constraints, including new multiply imaged systems
and redshifts.  The eight models shown in black at the top are from
the ``parametric'' family, and the six in green at the bottom are
``free-form.''  The grey dashed line marks the unweighted mean for the
set of all models, at $\mu=2.53$.
\label{fig:lensingtest} }
\end{center}
\end{figure}

@@

In Figure~\ref{fig:lensingtest} the model predictions are plotted
alongside the observed magnification of SN \tomas, derived in
Section~\ref{sec:Magnification}.  This comparison shows that these 14
models are largely consistent with each other. Naively treating each
model as an independent prediction for the magnification (and ignoring
their quoted uncertainties), one would get a mean for the full sample
of \change{$\mu=2.53\pm0.38$.  This is separated from the observed SN
magnification by $\delta\mu/\mu=25\%$, which is just over a
1$\sigma$ difference.}

@@

All of the models presented here use a fixed assumption for the
parameters of a standard \LCDM\ cosmology (typically: \Om=0.3,\OL=0.7,\Ho=70 km
s$^{-1}$ Mpc$^{-1}$).  The parameter values selected for this fiducial
cosmology do have an impact on the inferred magnifications, and
variation of these parameters must be accounted for when using cluster
mass models for cosmological tests \citep{Zitrin:2014b}. 
\citep{Bayliss:2015} demonstrated that
changing the cosmological parameter within the range of current
acceptable cosmologies \citep[e.g.][]{Hinshaw:2013,Planck:2015},
introduces a magnification uncertainty that is comparable in magnitude
to that of the statistical uncertainties.  This leads to the
conclusion that the lens modeling uncertainties presented here are
underestimated, and should be significantly increased.

@@

\subsection{Foreground Dust}
\label{sec:ForegroundDust}

@@

@@

J.M.D acknowledges support of the consolider project CAD2010-00064 and
AYA2012-39475-C02-01 funded by the Ministerio de Economia y
Competitividad.

>>>>

\begin{figure*}
\begin{center}
%\includegraphics[width=\textwidth]{FIG/discovery_image}
\includegraphics[width=\textwidth]{FIG/discovery_image_lowres}
\caption{  \label{fig:DiscoveryImage} 
SN \tomas\ in the Abell 2744 field.  The left panel shows a
UV/Optical/IR color composite image constructed from all available HST
imaging of the Abell 2744 cluster field.  The inset panels on the right
show F814W imaging of the immediate vicinity of SN \tomas,
approximately 40\arcsec\ from the center of the cluster. The top panel
shows the template image, combining all data prior to the SN
appearance.  Labeled ellipses mark the nearest galaxies and their
\changeG{spectroscopic} redshift constraints, with the most likely host
galaxy marked in blue, and a background galaxy in yellow. The bottom
panel is constructed from all HFF F814W imaging taken while the SN was
detectable, and marks the SN location with an arrow.  (Left panel
image credit: NASA, ESA, and J. Lotz, M. Mountain, A. Koekemoer, and
the HFF Team (STScI)) }
\end{center}
\end{figure*}

@@

To refine the redshift and phase constraints on \tomas, we next fit
the spectrum with a custom spectral matching program that employs a
library of Type Ia SN SEDs.  This library is similar to the Type Ia
spectral set used by SNID, but also includes more recent SNe with
well-observed spectral time series that extend to rest-frame UV
wavelengths (e.g. SN 2011fe and 2014J).  We first use an approach
similar to the SNID algorithm: warping the pseudo-continuum of each
template spectrum by dividing out a 3$^{\rm rd}$-order polynomial to
match the observed SED of SN \tomas.  \changeG{This approach will find
templates that have similar abundances and photospheric
velocities. With the redshift fixed at $z=1.3457$ we find the best fit
is a spectrum from the normal Type Ia SN 2014J \citep{Foley:2014},
with a $\chi^2$ per degree of freedom $\nu$ of $\chi^2/\nu=121.8/89$
shown in the top panel of Figure~\ref{fig:SpecFit}.  Allowing the
redshift as an additional free parameter, we still find results that
are consistent with the SNID fits and $<2\sigma$ from the host galaxy
redshift: $z=1.31\pm0.02$ and a phase of $0\pm3$ rest-frame days. The
best-fitting spectral template in this case is the normal Type Ia SN
2012cg \citep{Amanullah:2015} at $z=1.31$ with $\chi^2/\nu=98.5/88$
(second panel of Figure~\ref{fig:SpecFit}).}

\changeG{
Next, we repeat the fitting, but without any warping of the templates
to account for differences in the continuum shape.  In this iteration
we only allow each template SED to be scaled in flux coherently at all
wavelengths, so the fits are more sensitive to the overall color of
the SED. Fixing the redshift to $z=1.3457$, we find the best match is
from the normal Type Ia SN 2011fe \citep{Mazzali:2014}, though the fit is quite poor, with
$\chi^2/\nu=313.2/91$ (third panel of Figure~\ref{fig:SpecFit}).  When
the redshift is allowed as a free parameter the \tomas\ SED is still
matched best by a SN 2011fe template, now at redshift $z=1.31$ with
$\chi^2/\nu=240.8/90$ (fourth panel of Figure~\ref{fig:SpecFit}).  }

Without the continuum warping, an alternative fit also arises: the
fast-declining (91bg-like) Type Ia SN 2011iv \citep{Foley:2012b} at
$z=0.98\pm0.01$. Formally, this match provides a slightly better fit
to the unwarped \tomas\ spectrum \changeG{($\chi^2/\nu=220.6/90$,
bottom panel of Figure~\ref{fig:SpecFit}),} although the fit is
notably poorer at $\sim8700$\AA\ where the most significant absorption
feature is found. Furthermore, a redshift $z\sim1$ is at odds with the
spectroscopic redshift of the nearest galaxy ($z=1.3457$), and we will
see in the following section that the photometric data is also
incompatible with a Type Ia SN at $z\sim1.0$.

@@

\section{Distance Modulus and Magnification}
\label{sec:DistanceAndMagnification}

With the type and redshift securely defined as a normal Type Ia SN at
$z=1.3457$, we now turn to fitting the light curve with Type Ia
templates to measure the distance modulus \changeG{ and then derive
the gravitational lensing magnification.  In this process, we would
like to avoid introducing systematic uncertainties inherent to any
assumed cosmological model \citep[e.g.][]{Nordin:2014}.  To that end,
we follow \citealt{Patel:2014} and define the magnification by
comparing the measured distance modulus of SN \tomas\ against an
average distance modulus derived from a ``control sample'' of unlensed
Type Ia SNe at similar redshift.  This allows us to make only the
minimal assumption that the redshift-distance relationship for Type Ia
SN is smooth and approximately linear over a small redshift span,
which should be true for any plausible cosmological model.  
}

@@

\subsection{Light Curve Fitting}
\label{sec:LightCurveFitting}

\changeG{
As in \citealt{Patel:2014}, we derive a distance modulus for
SN \tomas\ and all SNe in our control sample using two independent
light curve fitters: the SALT2 model described above and the MLCS2k2
model \citep{Jha:2007}.  }  With both fitters we find light curve
shape and color parameters for \tomas\ that are fully consistent with
a normal Type Ia SN.  For SALT2, with the
\change{redshift fixed at $z=1.3457$, we find a light curve shape
parameter of $x_1=0.135\pm0.199$ and a color parameter of
$c=-0.127\pm0.025$, yielding a $\chi^2$ value of 45.0 for 36 degrees
of freedom, $\nu$.  With the MLCS2k2 fitter the best-fit shape
parameter is $\Delta=-0.082\pm0.070$ and the color term is
$A_{\rm V}=0.011\pm0.025$, giving $\chi^2/\nu=22.8/36$.}

\changeG{
 The MLCS2k2 fitter returns a distance
modulus\footnote{We use 'dm' to indicate the distance modulus to avoid
confusion, reserving the symbol $\mu$ to refer to the lensing
magnification. This 'dm' is a standard distance modulus, defined as
${\rm dm}=5\log_{10}d_L+25$, where $d_L$ is the luminosity distance in
Mpc.}  $\rm{dm}_{\rm MLCS2k2}$ directly, as it is defined to be one of
the free parameters in the model.} To derive dm from the SALT2 fit, we
use

@@

Final values for the SN distance modulus are shown in
Table~\ref{tab:MeasuredMagnification}, adopting the spectroscopic
redshift ($z=1.3457$).  In addition to modifying the distance modulus
uncertainty for SN \tomas, we also add $\sigma_{\rm int}=0.08$ mag in
quadrature to the uncertainty for every SN in the control sample. Note
however that the uncertainty on the control sample value at $z=1.3457$
is only 0.06 mag, smaller than the intrinsic dispersion \changeG{of
any single SN} because it reflects our measurement error on the mean
distance modulus of the population.  The distance moduli derived from
the SALT2 and MLCS2k2 light curve fitters are fully consistent within
the uncertainties.

\subsubsection{Host Galaxy Mass Correction}
\label{sec:HostGalaxyMassCorrection}

@@

The physical mechanism that drives the mass step is not yet
understood, but may be related to the metallicity or age of the SN
progenitor systems.  In either case, the significance of this effect
should decrease with redshift, as metal-rich passive galaxies become
much less common at $z>1$ \change{\citep[see
e.g.][]{Rigault:2013,Childress:2014b}.}  Indeed, when the size of the
mass step correction is allowed to vary with redshift, there is no
significant evidence that a non-zero mass step is required at
$z>1$ \citep{Suzuki:2012,Shafer:2014,Betoule:2014}.  Given the
absence of a clear physical model and the lack of empirical support
for a high-$z$ mass step, we do not apply any correction to SN \tomas\
or the control sample.

\subsection{Control Sample Comparison}
\label{sec:ControlSampleComparison}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{FIG/snTomas_magnifications}
\caption{ 
Comparison of the observed lensing magnification to predictions from
lens models. The vertical blue line shows the constraints from
SN \tomas\ derived in Section~\ref{sec:DistanceAndMagnification} using the
MLCS2k2 fitter, with a shaded region marking the total
uncertainty. Markers with horizontal error bars show the median
magnification and 68\%\ confidence region from each of the 14 lensing
models.  Circles indicate models that use only strong-lensing
constraints, while diamonds denote those that also incorporate
weak-lensing measurements.  Models shown with open markers were
constructed using only data available before the start of the Frontier
Fields observations.  Filled markers indicate models that use
additional input constraints, including new multiply imaged systems
and redshifts.  The eight models shown in black at the top are from
the ``parametric'' family, and the six in green at the bottom are
``free-form.''  The grey dashed line marks the unweighted mean for the
set of all models, at $\mu=2.53$.
\label{fig:lensingtest} }
\end{center}
\end{figure}

@@

In Figure~\ref{fig:lensingtest} the model predictions are plotted
alongside the observed magnification of SN \tomas, derived in
Section~\ref{sec:DistanceAndMagnification}.  This comparison shows that these 14
models are largely consistent with each other. Naively treating each
model as an independent prediction for the magnification (and ignoring
their quoted uncertainties), one would get a mean for the full sample
of \change{$\mu=2.53\pm0.38$.  This is separated from the observed SN
magnification by $\delta\mu/\mu=25\%$, which is just over a
1$\sigma$ difference.}

@@

All of the models presented here use a fixed assumption for the
parameters of a standard \LCDM\ cosmology (typically: \Om=0.3,\OL=0.7,\Ho=70 km
s$^{-1}$ Mpc$^{-1}$).  The parameter values selected for this fiducial
cosmology do have an impact on the inferred magnifications, and
variation of these parameters must be accounted for when using cluster
mass models for cosmological tests \citep{Zitrin:2014b}. 
\citep{Bayliss:2015} demonstrated that
changing the cosmological parameter within the range of current
acceptable cosmologies \citep[e.g.][]{Hinshaw:2013,Planck-Collaboration:2015},
introduces a magnification uncertainty that is comparable in magnitude
to that of the statistical uncertainties.  This leads to the
conclusion that the lens modeling uncertainties presented here are
underestimated, and should be significantly increased.

@@

\subsection{Foreground Dust and SN Color}
\label{sec:ForegroundDust}

@@

\changeG{
Although the very blue color of SN \tomas\ is helpful to rule out dust
as a possible explanation for the observed small systematic bias, it
is possible that this very blue color is itself leading to a bias in
the distance measurement.  \citet{Scolnic:2014b} measured a small bias
in the SALT2 fitter for Type Ia SNe that have a color parameter
derived from the light curve fit $c<-0.1$ (see, e.g., their Figure 10
and Section 5.1). For \tomas\ we have measured $c=-0.13$, which would
correspond to a bias of roughly $+0.05$ mag in the SALT2 distance
measurement.  If we were to apply a $-0.05$ mag correction to the
distance modulus from the SALT2 fit, then this would increase the
inferred magnification to $\mu_{\rm SALT2}=2.08\pm0.36$.  This would
still be consistent with the mild tension apparent from the MLCS2k2
fit, though with a decrease in significance. 
}

@@

J.M.D acknowledges support of the consolider project CSD2010-00064 and
AYA2012-39475-C02-01 funded by the Ministerio de Economia y
Competitividad.
